### 2025-08-09 basic-statistics.qmd
20250809 - SimDesign simulation

```{r}
library("petersenlab")
library("DescTools")
library("pwr")
library("pwrss")
library("WebPower")
library("grid")
library("psych")
library("tidyverse")
```

@@

The power analysis code above assumes the groups are of equal size (i.e., a balanced design).
If the design is unbalanced (i.e., there are different numbers of participants in each group), it may be necessary to conduct a power analysis via a simulation.
Below is an example of evaluating the statistical power for detecting an effect unbalanced designs via simulation:

```{r}
nSim <- 1000 # number of simulations

# Function to generate data and perform ANOVA
simulate_anova <- function(nGroup1, nGroup2, f, alpha) {
  # Means for each group
  mean1 <- 0
  mean2 <- f * sqrt((nGroup1 + nGroup2) / 2)

@@

  # Combine data
  data <- data.frame(
    value = c(group1, group2),
    group = factor(rep(c("Group1", "Group2"), c(nGroup1, nGroup2)))
  )

  # Perform ANOVA
  aov_result <- aov(value ~ group, data = data)
  p_value <- summary(aov_result)[[1]][["Pr(>F)"]][1]

  # Check if p-value is less than alpha
  return(p_value < alpha)
}

# Run simulations
set.seed(52242) # for reproducibility
powerSimulationOneWayAnova <- replicate(
  nSim,
  simulate_anova(
    nGroup1 = 10,
    nGroup2 = 25,
    f = effectSize_f,
    alpha = alpha))

# Estimate power
mean(powerSimulationOneWayAnova)
```

>>>>

```{r}
library("petersenlab")
library("DescTools")
library("pwr")
library("pwrss")
library("WebPower")
library("grid")
library("psych")
library("SimDesign")
library("tidyverse")
```

@@

##### Simulation {#sec-statisticalPowerOneWayANOVAsimulation}

The power analysis code above assumes the groups are of equal size (i.e., a balanced design).
If the design is unbalanced (i.e., there are different numbers of participants in each group), it may be necessary to conduct a power analysis via a simulation.
Below is an example of evaluating the statistical power for detecting an effect unbalanced designs via simulation using the `SimDesign` [@R-SimDesign; @Chalmers2020_packages] package.
We discuss simulation more in @sec-simulation.

###### 1. Design Matrix {#sec-statisticalPowerOneWayANOVAsimulationDesign}

First, we set up the design matrix using the `SimDesign::createDesign()` function:

```{r}
Design <- SimDesign::createDesign(
  nGroup1 = c(10),
  nGroup2 = c(25, 50, 100),
  f       = c(.10, .25, .40), # small, medium, large effect sizes
  alpha   = 0.05
)
```

###### 2. Generate {#sec-statisticalPowerOneWayANOVAsimulationGenerate}

Second, we set up the `Generate()` function:

```{r}
Generate <- function(condition, fixed_objects = NULL) {
  with(condition, {
    # Means for each group
    mean1 <- 0
    mean2 <- f * sqrt((nGroup1 + nGroup2) / 2)

@@

    data.frame(
      value = c(group1, group2),
      group = factor(rep(c("Group1", "Group2"), 
                         c(nGroup1, nGroup2)))
    )
  })
}
```

###### 3. Analyse {#sec-statisticalPowerOneWayANOVAsimulationAnalyse}

Third, we set up the `Analyse()` function:

```{r}
Analyse <- function(condition, dat, fixed_objects = NULL) {
  aov_result <- aov(value ~ group, data = dat)
  p_value <- summary(aov_result)[[1]][["Pr(>F)"]][1]

  data.frame(
    p_value  = p_value,
    rejectH0 = as.numeric(p_value < condition$alpha)
  )
}
```

###### 4. Summarise {#sec-statisticalPowerOneWayANOVAsimulationSummarize}

Fourth, we set up the `Summarise()` function:

```{r}
Summarise <- function(condition, results, fixed_objects = NULL) {
  # Estimated power = proportion of rejections
  Power <- mean(results$rejectH0)

  data.frame(Power = Power)
}
```

###### 5. Run the Simulation {#sec-statisticalPowerOneWayANOVAsimulationRun}

Fifth, we run the simulation using the `SimDesign::runSimulation()` function:

```{r}
monteCarloSim_results <- SimDesign::runSimulation(
  design       = Design,
  replications = 1000,
  generate     = Generate,
  analyse      = Analyse,
  summarise    = Summarise,
  seed = genSeeds(Design, iseed = 52242), # for reproducibility
  parallel = TRUE # for faster (parallel) processing
)
```

###### Simulation Results {#sec-statisticalPowerOneWayANOVAsimulationResults}

Here are the simulation results:

```{r}
monteCarloSim_results
```
